---
title: "Homework 1"
subtitle: "Computational Methods for American Politics"
author: "Sanittawan Tan"
date: "10/22/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load Libraries

```{r libraries}
library(ggplot2)
library(tidyverse)
```


## Part 1 - Load data set

```{r load data}
dir <- "/Users/sanittawan/Documents/Nikki/UChicago/Classes/Autumn_2019/com_methods_am_politics/psets/problem-set-1/State Leg Prof Data & Codebook/legprof-components.v1.0.RData"
load(dir)
SLPdta <- as_tibble(x)
```

## Part 2 - Munge the data

### a) Select only continuous features

```{r filter continuous feats}
SLPdta_filtered <- SLPdta %>% 
  select(-mds1, -mds2)
```


### b) Restrict data to 2009/10 legislative session

```{r select sessid 2009/10}
SLPdta_filtered2 <- SLPdta_filtered %>% 
    filter(sessid == "2009/10")
```

### c) Omit missing values

```{r deal with NAs}
SLPdta_filtered3 <- SLPdta_filtered2 %>%
    select_if(~ !any(is.na(.)))
```


### d) Standardize input features

```{r}

```


### e) Miscellaneous

```{r}
state_names <- unique(SLPdta_filtered$state) 
state_abbr <- unique(SLPdta_filtered$stateabv)
```


## Part 3 - Diagnose Clusterability


## Part 4 - Agglomerative Hierachical Clustering


## Part 5 - k-Means 


## Part 6 - Gaussian Mixture Model (GMM)


## Part 7 - Plotting


## Part 8 - Validation Strategy


## Part 9 - Discussion on Validation Output

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
